---
title: "Practical Machine Learning - Course Project"
author: "Guillermo Ibarra"
date: "April 16, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 

More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

The goal of this project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. The prediction model will then be used to predict 20 different test cases.

```{r setup_eviron, echo=TRUE,warning=FALSE,error=FALSE,cache=TRUE,message=FALSE}
# load packages
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
library(knitr)
library(doParallel)

# Clean up environment
#rm(list=ls())
#gc()
```

## Get data
The training data for this project are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

To load the datasets we execute:

```{r load_data, echo=TRUE,warning=FALSE,error=FALSE,cache=TRUE,message=FALSE}
# download data or use previously downloaded data
if(!file.exists("pml-training.csv")){
  download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", 
                destfile = "pml-training.csv", method = "curl")
}
if(!file.exists("pml-testing.csv")){
  download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", 
                destfile = "pml-testing.csv", method = "curl")
}
```

# Load data with valid values
```{r clean_data, echo=TRUE,warning=FALSE,error=FALSE,cache=TRUE,message=FALSE}
train <- read.csv("pml-training.csv", header = TRUE, na.strings=c("","NA", "#DIV/0!"))
test <- read.csv("pml-testing.csv", header = TRUE, na.strings=c("","NA", "#DIV/0!"))
```

## Check valid data

```{r how_clean, echo=TRUE,warning=FALSE,error=FALSE,cache=TRUE,message=FALSE}
# see error percentage 
NAPercent <- round(colMeans(is.na(train)), 2)
table(NAPercent)

```

## Clean up variables

```{r clean_vars, echo=TRUE,warning=FALSE,error=FALSE,cache=TRUE,message=FALSE}
# find index of the complete columns minus the first 
index <- which(NAPercent==0)[-1]
# subset the data
train <- train[, index]
test <- test[, index]
# looking at the structure of the data for the first 10 columns
str(train[, 1:10])


# get subset the data
train <- train[, -(1:6)]
test <- test[, -(1:6)]
# convert all numerical data to numeric
for(i in 1:(length(train)-1)){
  train[,i] <- as.numeric(train[,i])
  test[,i] <- as.numeric(test[,i])
}
```

## Cross validation

We split the data, 80% goes to the training set, 20% goes to the testing set.

```{r cross_val, echo=TRUE,warning=FALSE,error=FALSE,cache=TRUE,message=FALSE}
# split train data set
inTrain <- createDataPartition(y=train$classe,p=0.8, list=FALSE)
trainData <- train[inTrain,]
validation <- train[-inTrain,]
# print out the dimensions of the 3 data sets
rbind(trainData = dim(trainData), validation = dim(validation), test = dim(test))
```

## Modeling

We model with Generalized Boosted Regression Model

```{r mod_gbrm, echo=TRUE,warning=FALSE,error=FALSE,cache=TRUE,message=FALSE}
# run the generalized boosted regression model
gbmFit <- train(classe~., data = trainData, method ="gbm", verbose = FALSE)
gbmFit
# use model to predict on validation data set
gbmPred <- predict(gbmFit, validation)
# predicted result
confusionMatrix(gbmPred, validation$classe)
```

Then we model with Random Forests

```{r mod_rf, echo=TRUE,warning=FALSE,error=FALSE,cache=TRUE,message=FALSE}
# run the random forest model on the training data set
rfFit <- randomForest(classe~., data = trainData, method ="rf", prox = TRUE)
rfFit
# use model to predict on validation data set
rfPred <- predict(rfFit, validation)
# predicted result
confusionMatrix(rfPred, validation$classe)
```
## Comparison of models
The better performing algorithm is **Random forest**  with a **0.45%** error rate. The cross validation yields that the model achieves an accuracy of **99.6%**, while the  **Generalized Boost Model** has an accuracy of **96.0%**.

## Result
This is the result for the 20 test cases.
```{r results_qz, echo=TRUE,warning=FALSE,error=FALSE,cache=TRUE,message=FALSE}
# apply random forest model to test set
predict(rfFit, test)

```